%!TEX root=000-main.tex

\chapter{Global States}
\label{chap:global_states}
Global states (or flags) are defined at datapath level and are not related to a single flow of a particular stage.
Flags are 32 boolean registers and are grouped in an uint32\_t variable global\_states in the datapath struct.

\section{Flags match field}
\label{section:oxm_of_flags}

If a switch supports OpenState (flag OFPTC\_TABLE\_STATEFUL set), right after the packet headers are parsed, the global states are retrieved and written in the flags field. OXM\_OF\_FLAGS is a field with mask, so it is possible to match it either exactly or with wildcards. A 0 bit in the mask means i-th flags value is ``do not care'', while a 1 bit value means ``exact match''.

Definition in ofsoftswitch13 (oxm-match.h file)
\begin{verbatim}
/* Global States */
#define OXM_OF_FLAGS OXM_HEADER     (0x8000, 40, 4)
#define OXM_OF_FLAGS_W OXM_HEADER_W (0x8000, 40, 4)
\end{verbatim}
Example match:
\begin{verbatim}
flags=(4,5)
\end{verbatim}
This command allows to match over *****************************1*0 flags configuration (4 in binary is 100 and the mask 5 is 101 that is exact match on LSB 1 (0 value) and LSB 3 (1 value) and ``don’t care'' over all the other flags. In order to perform an exact match on flags value no mask is required.
\\Example match:
\begin{verbatim}
flags=4
\end{verbatim}
NB: this match is very differend from the previous one. With this command we are matching over 00000000000000000000000000000100 flags configuration, so it is an exact match.

\section{Set flag action}
\label{section:set_flag_action}
The OFPAT\_SET\_FLAG action is used to set flags' value.
In the flow modification message, the controller defines the action with the following parameters:
\begin{itemize}
\setlength\itemsep{0em}
\item flag = flags value
\item mask = flags mask
\end{itemize}

\paragraph{Structures in ofsoftswitch13:}\mbox{}\\
\begin{verbatim}
OFPAT_SET_FLAG = 29,   /* Set a single flag value of the global state */

/* Action structure for OFPAT_SET_FLAG */
struct ofp_action_set_flag {
    uint16_t type; /* OFPAT_SET_FLAG */
    uint16_t len;  /* Length is 8. */
    uint32_t value; /* flag value */
    uint32_t mask;    /*flag mask*/
    uint8_t pad[4];   /* Align to 64-bits. */
};
OFP_ASSERT(sizeof(struct ofp_action_set_flag) == 16);
\end{verbatim}

\section{Flag modification messages}
\label{section:flag_mod_msg}
The OFPT\_FLAG\_MOD message allow the controller to modify global states value.
\paragraph{Structures in ofsoftswitch13:}\mbox{}\\
\begin{verbatim}
OFPT_FLAG_MOD = 31,  /* Controller/switch message */

struct ofp_flag_mod {
    struct ofp_header header;
    uint32_t flag;
    uint32_t flag_mask;
    uint8_t command;
    uint8_t pad[7];                  /* Pad to 64 bits. */
};

enum ofp_flag_mod_command { 
    OFPSC_MODIFY_FLAGS = 0,
    OFPSC_RESET_FLAGS
};

\end{verbatim}

There are two message types.
\subsection{Modify flags command}
The OFPSC\_MODIFY\_FLAGS allows to modify global states values. The controller sends a OFPSC\_MODIFY\_FLAGS message with the following parameters:
\begin{itemize}
\setlength\itemsep{0em}
\item flag = flags value
\item flag\_mask = mask value
\item command = 0 (OFPSC\_MODIFY\_FLAGS)
\end{itemize}

\subsection{Reset flags command}
The OFPSC\_MODIFY\_FLAGS allows to reset global states to the default value (OFP\_GLOBAL\_STATES\_DEFAULT in openflow.h)
The controller sends a OFPSC\_MODIFY\_FLAGS message with the following parameters:
\begin{itemize}
\setlength\itemsep{0em}
\item flag = ANY
\item flag\_mask = ANY
\item command = 1 (OFPSC\_RESET\_FLAGS)
\end{itemize}